---
title: "SFBay native oyster survival analyses"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Karina J. Nielsen"
date: "Last compiled on `r format(Sys.time(), '%d %B %Y')`"
output:
   bookdown::html_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  fig.align = "left",
  fig.width = 8,
  fig.height = 11,
  dev = "png",
  cache = T)

# echo = FALSE means code will not print, by default
# cache = TRUE saves output results of all chunks so that they do not need to be rerun

```

# Question of interest
Do juvenile native oysters in San Francisco Bay survive better when they have rockweed <i> Fucus distichus </i> cover (+/-) Does the effect, if any, vary with tidal height (low vs. high)? 

# Methods summary
Juvenile <i> Ostrea lurida</i> were super-glued to individually labeled porcelain tiles and attached to rocks in the intertidal zone with marine epoxy putty. We established ten plots of ten oysters each with and without <i> Fucus </i> cover in the high and low intertidal zones.

Oyster plots were established on January 19, 2020. We surveyed the plots every two weeks until May 2, 2020 when we terminated the experiment for all but two plots we inadvertently missed. We terminated these two plots on June 11, 2020. We missed one survey in mid-March due to the pandemic stay at home health order. 

For each survey, we noted the condition of each oyster and photographed the plots. We noted whether or not the tile, label, or oyster was missing. We also noted if the oyster was alive or if it was dead with some shell remaining.

Data were entered in an Excel spreadsheet and then tidied and analyzed using R Statistical Software version 4.0X, R Core Team (2020). Time intervals were calculated to determine days elapsed from the beginning of deployment to the last follow-up time for each individual oyster. The data were right-censored unless a dead oyster (shell present) was observed. 

```{r load packages using easypackages, include=FALSE}
#renv::init() # to initialize a new project-local environment with a private R library

#renv::restore()

#reference for renv workflow: https://rstudio.github.io/renv/articles/renv.html
my_packages <- c("renv", "easypackages",  "here", "ggthemes", "gridExtra", "tidyverse", "lubridate", "patchwork", "readxl", "survival", "survminer", "ranger", "condSURV", "ggplot2", "gtsummary", "flextable", "dplyr", "forcats","bookdown") 
# list of packages to be used in this Rmd file
easypackages::packages(my_packages) 
# checks to make sure the packages needed are loaded and asks before installing to confirm install in the console window
easypackages::libraries(my_packages) # could be used in lieu of packages() if you already have the packages installed

```


```{r Confirm working directory for project, include = FALSE}
here::here() # use this to work with sub-directories less awkwardly and share projects across users/machines more easily.
```


```{r renv, include=FALSE}
#Run this to update snapshot of packages when you commit the file
#renv::snapshot() 
#run as needed later in the project to capture required package versions
#NB must respond and affirms actions in console or you will see the spinning wheel....

```


```{r read in data, include=FALSE}

#read in data from xlsx file sheet
oyster_survival <- read_excel(here::here("data", "Oyster Survival Experiment Data Entry aug2020.xlsx"), sheet = "data", col_names = TRUE)


```

# Survival analyses
We used the survival package Therneau T (2020) to calculate survival curves with the Kaplan-Meier method. A log-rank chi-squared test was used to assess differences among the curves. We also fit Cox proportional hazards regression model.
<br>

```{r wranglin, include=FALSE}


# defining treatment levels, ordering
oyster_survival$zone <- factor(oyster_survival$zone, levels = c("L", "H"))
oyster_survival$fucus <- factor(oyster_survival$fucus, levels = c("F", "NF"))
oyster_survival$zone <- fct_recode(oyster_survival$zone, low = "L", high = "H")
oyster_survival$fucus <- fct_recode(oyster_survival$fucus, bare = "NF", fucus = "F")

# Calculate days of oyster survival in new variable lifetime
# end date is last day the oyster was observed, not the day it was missing
oyster_survival <-  mutate(oyster_survival, lifetime = as.numeric(difftime(oyster_survival$end, oyster_survival$deploy, units = "days")))



```

```{r box_plot_wranglin, include=FALSE}
# reworking plot name
# reclassified as character for categorical analysis
# renamed to avoid conflating with ggplot
oyster_survival$plot <- as.character(oyster_survival$plot)
oyster_survival <-  rename(oyster_survival, oyst_plot = plot)

# create shade variable to call out shaded plots
shady_list <- c(41, 42, 38, 37, 35, 33, 29)

oyster_survival <- mutate(oyster_survival,
                          shade = case_when(
  oyst_plot == shady_list[1] ~ 1,
  oyst_plot == shady_list[2] ~ 1,
  oyst_plot == shady_list[3] ~ 1,
  oyst_plot == shady_list[4] ~ 1,
  oyst_plot == shady_list[5] ~ 1,
  oyst_plot == shady_list[6] ~ 1,
  oyst_plot == shady_list[7] ~ 1,
  TRUE ~ 0
)
)

# create sun variable to call out sun exposed plots 
sunny_list <- c(21, 22)

oyster_survival <- mutate(oyster_survival,
                          sun = case_when(
  oyst_plot == sunny_list[1] ~ 1,
  oyst_plot == sunny_list[2] ~ 1,
  TRUE ~ 0
)
)

# create low cover variable to call out plots that lost fucus cover

low_cover_list <- c(14, 29, 39, 40) #plot 40 too but was removed from data since all tiles fell off

oyster_survival <- mutate(oyster_survival,
                          low_cover = case_when(
  oyst_plot == low_cover_list[1] ~ 1,
  oyst_plot == low_cover_list[2] ~ 1,
  oyst_plot == low_cover_list[3] ~ 1,
  TRUE ~ 0
)
)

# filtered data frame for box plot
oyster_survival_box <- oyster_survival

# dropping oysters where the whole tile, oyster or label was missing, coded C,A or B status unknown 
oyster_survival_box  <- mutate(filter(oyster_survival_box, status != "C")) 
oyster_survival_box  <- mutate(filter(oyster_survival_box, status != "B"))
oyster_survival_box  <- mutate(filter(oyster_survival_box, status != "A"))

# create dummy variable for surviving oysters
oyster_survival_box <- mutate(oyster_survival_box, present = if_else(status == "Z", 1, 0))

# creating treatment variable
oyster_survival_box <- mutate(oyster_survival_box, treatment = paste0(zone,"+",fucus))

#### BOX PLOT DATA WRANGLIN ####

#total oysters in plots. missing tiles and oysters already removed.

tot_by_plot_all <- oyster_survival_box %>%
  group_by(oyst_plot) %>%
  summarise(n_tot = n(),
            trt = first(treatment),
            shade = first(shade),
            sun = first(sun),
            low_cover = first(low_cover)
            )

# total survivors remaining. missing tiles and oysters already removed
tot_plot_surv <- oyster_survival_box %>% 
  filter(status == "Z") %>%
  group_by(oyst_plot) %>%
  summarise(n_surv = n(), )


box_data <- full_join(tot_by_plot_all, tot_plot_surv, by = "oyst_plot")

box_data$n_surv <- replace_na(box_data$n_surv, 0)

box_data <- mutate(box_data, pct_surv = round((n_surv/n_tot)*100, digits = 1) )

#### Box plot DATA WRANGLIN ####

#total oysters in treatment. missing tiles and oysters already removed.

tot_by_trt <- oyster_survival_box %>% 
  group_by(treatment) %>% 
  summarise(n_tot = n(), )

# total survivors remaining. missing tiles and oysters already removed
tot_surv <- oyster_survival_box %>% 
  filter(status == "Z")

#total survivors by treatment
tot_surv <- tot_surv %>% 
  group_by(treatment) %>% 
  summarise(n_surv = n())

# #summary statistics
# # using method from https://www.dummies.com/education/math/statistics/how-to-calculate-the-margin-of-error-for-a-sample-proportion/
# 
# bar_data <- bind_cols(tot_by_trt,tot_surv)
# 
# bar_data <- rename(bar_data, 
#                    treatment = treatment...1,
#                    treatments = treatment...3,
#                    )
# 
# bar_data <- bar_data %>% 
#   mutate(surv_propotion = n_surv/n_tot,
#          one_minus_surv_p = 1-surv_propotion,
#          margin_error_surv_p = sqrt((surv_propotion*one_minus_surv_p/n_tot)),
#          surv_pct = surv_propotion * 100,
#          surv_mrg_err_pct = margin_error_surv_p*100
#          )
# 

```

```{r censoring, include = FALSE}

oyster_survival <-  rename(oyster_survival, field_stat = status)

oyster_survival <- mutate(oyster_survival, status = if_else(field_stat == "D", 1, 0))

```

## Kaplan-Meier survivorship analysis

```{r km_model, include=FALSE}
# see: https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html
# see: https://raw.githubusercontent.com/rstudio/cheatsheets/master/survminer.pdf

#K-M

#fit_a <- surv_fit(Surv(lifetime, status) ~ zone + fucus, data = oyster_survival)
#table <- ggsurvtable(fit_a, data = oyster_survival, color = "strata", y.text = FALSE)

# summary(fit_a) # use this to get table of survivorship estimates at every survey point

# survdiff(Surv(lifetime, status) ~ zone + fucus, data = oyster_survival) # chi2 model// V = theoretical variance in O-E, assumed to be chi2


```


```{r s_plot}
ggsurvplot(
    fit = fit_a,
    xlab = "Days",
    ylab = "Overall survival probability (%)",
    censor = FALSE,
    pval = TRUE,
    fun = "pct",
    risk.table = TRUE,
    risk.table.y.text.col = T,
    risk.table.y.text = FALSE,
    ncensor.plot = TRUE,
    ggtheme = theme_minimal(),
    size = 1.5,
    palette = c("steelblue4","steelblue1", "darkorange4", "darkorange"),
    legend = "top",
    legend.title = "",
    legend.labs = c("low zone + fucus", "low zone", "high zone + fucus", "high zone")
)
```

## Cox proportional hazards analysis
### Table

```{r cox_model, message=FALSE}

# Cox

f2 <- coxph(Surv(lifetime, status) ~ zone * fucus, data = oyster_survival)
#summary(f2)
# ggcoxdiagnostics(f2, type = "deviance")
f2 %>% 
  gtsummary::tbl_regression(exp = TRUE) %>% 
  add_global_p()%>% 
  bold_labels() %>% 
  italicize_levels()%>%
  as_flex_table()

```

### Forest plot
```{r forest_plot}
for.plot <- ggforest(f2, data = oyster_survival, main = "Hazard Ratio",   cpositions = c(0.01, 0.1, 0.26), fontsize = 1)
for.plot
# ggsave("forest.png", width = 8, height = 8, units = "in", dpi="print")
```

## Oysters with known fates - proportion surviving

For all oysters that had a known fate at the end of the field experiment (i.e., persisted alive until the end, or died, leaving behind some shell on the tile), we calculated the proportion of these that survived in each plot for each of the four unique treatment combinations (zone (low/high) x fucus (+/-)). We present these as box plots. We also indicate the plots in the high zone that appeared to benefit from some shading from adjacent trees at the edge of the upper intertidal zone.

```{r last,   fig.width = 6,  fig.height = 6,}

# last_ones <- ggplot(box_data, aes(x=trt, y=pct_surv, fill=trt)) +
#    geom_boxplot(alpha=0.3, outlier.shape = NA) +
#   geom_point(pch = 21,
#              position = position_jitterdodge(
#                jitter.height = 2,
#                jitter.width = 0.5,
#                seed = 7
#              )) +
#   theme_few() +
#   theme(legend.position = "none") +
#   labs(x = "Treatment Groups\n High (H) or Low Tidal Zone (L)\n Fucus distichus Cover Present (F) or Absent (NF)",  y = "Percent of Oysters Survived per Replicate") +
#   ggtitle("Oyster Survival by Treatment")
# 
# last_ones

unshaded_df <- box_data %>%
  filter(shade < 1)

shaded_df <- box_data %>%
  filter(shade > 0)


last_ones_2 <- ggplot(box_data, aes(x=trt, y=pct_surv, fill = trt))+ 
 geom_boxplot(alpha=0.7, outlier.shape = NA) +
  geom_point(data = unshaded_df,
             pch = 21,
             alpha = 0.7,
             size = 3,
             stroke=1,
             position = position_jitterdodge(
               jitter.height = 1,
               jitter.width = 0.5,
               seed = 7)) +
  geom_point(data = shaded_df,
             aes(x = trt, y = pct_surv),
             color = 'black',
             pch = 19,
             size = 3,
             position = position_jitterdodge(
               jitter.height = 1,
               jitter.width = 0.5,
               seed = 7)) +
  scale_fill_manual(values=c("darkorange", "darkorange4", "steelblue1","steelblue4"))+
  scale_shape_manual(values = c(21,24)) +
  theme_few() +
  theme(legend.position = "none") +
  labs(x = "Treatment",  y = "percent surviving (per plot)") +
  ggtitle("Oyster survival (known fates only)") 

  
last_ones_2

```

